\Title

### 2013-02-10 synmut.tex
supplement only captions, more details about simulations. need to redo simulations without complex epitopes for 4B

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations. \comment{fix x-axis label}
Panel A) Time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red histogram) are systematically larger than of alleles are lost (blue
) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. The
average over all mutations that rare not observed  (green) lies between the
those that fix or get lost.
To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved streches between the variable loops, we consider mutations in
variable loops and conserved parts separately. The biggest depression in
fixation probability is observed in the conserved stems, while the variable
loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
is consistent with important stem structures in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the region V1-V5$\pm 50bp$)????. This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B) Fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. \comment{Can't we show this
for simple epitopes? I can't seem to find a way to describe this from the
outset}.
This simulation assumes 6 roughly equivalent escape mutations per epitope; see supplement for alternative models.}
\label{fig:simheat}
\end{center}
\end{figure}

@@

A functional significance of the insulating RNA structure stems between the
hyper-variable loops has been proposed
previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
Conserved RNA structures exist in different parts of the HIV genome (several are
known), but our analysis requires longitudinal data most of which focusses
on \env.
Furthermore, we only expect to see deleterious synonymous
mutations at high frequencies if they are linked to repeated adaptive
substitions. Otherwise, they remain at low frequencies and can only be observed
by deep sequencing methods.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence data collection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Longitudinal intrapatient viral RNA sequences were collected from published
studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
show substantial population structure and were discarded (see supplementary
materials); a total of 11 patients with 6-?? time points each and 10
sequences per time point were analyzed. The time intervals between two
consecutive sequences ranged from 6 to 18 months. \comment{CHECK VALUES}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimate of times to fixation/extinction, polymorphisms were
binned by frequency and the time to reaching the first boundary (fixation or
extinction) was stored. The fixation probability was determined as the long-time
limit of the resulting curves. Samples less than ???days
prior the last time point were omitted since there was not enough information
on their future fate, i.e., whether they fix or are lost.

The SHAPE scores quantifying the degree of base pairing of individuals sites in
the HIV genome were downloaded from the journal
website~\citep{watts_architecture_2009}.
Where possible, SHAPE reactivities were assigned to sites in the multiple
sequence alignments for each patient through the alignment to the sequence
of the NL4.3 virus used in \citep{watts_architecture_2009}. Problematic
assignments in indel-rich regions are excluded from the analysis.
\comment{The following is super confusing}
In order to restrict the analysis to synonymous polymorphisms, a lower frequency
threshold of 0.15 was used (other thresholds yield the same results). Since very
few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
enriched for to-be-lost mutations; hence the ``lost" curve in \FIG{SHAPEA}
contains much more points than the ``fixed" one.

The variable loops and flanking regions are identified manually starting from
the annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
similar approach is used to label the C2-V5 region sequenced in
ref.~\citep{shankarappa_consistent_1999}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Simulations are performed using FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
\env{}-only simulations are performed and yield comparable results. In each
run, a random fitness landscape with specified statistical properties 
(e.g.~density of beneficial sites, average deleterious effect of synonymous changes) is generated.
For each set of parameters, approximately 100 simulation runs are averaged over. 
\comment{needs a little bit more detail}.
For the discussion of simulation parameters, the areas below or above the neutral
diagonal were estimated from the binned fixation probabilities using linear
interpolation between the bin centers. This measure is sufficiently precise for
our purposes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Methods availability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All analysis scripts, as well as the sequence alignment used, are available for download.

>>>>

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) Time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red histogram) are systematically larger than of alleles are lost (blue
) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. The
average over all mutations that are not observed (green) lies between the
those that fix and those that get lost.
To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved streches between the variable loops, we consider mutations in
variable loops and conserved parts separately. The biggest depression in
fixation probability is observed in the conserved stems, while the variable
loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
is consistent with important stem structures in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the regions V1-V5$\pm 100bp$ from each). This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B) Fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. \comment{Can't we show this
for simple epitopes? I can't seem to find a way to describe this from the
outset}. \comment{FIXME: (i) only x acis is logarithmic; (ii) redo without
epitopes (less representative in the end, but easier to introduce).}
This simulation assumes 6 roughly equivalent escape mutations per epitope; see supplement for alternative models.}
\label{fig:simheat}
\end{center}
\end{figure}

@@

A functional significance of the insulating RNA structure stems between the
hyper-variable loops has been proposed
previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
Conserved RNA structures exist in different parts of the HIV genome (several are
known), but our analysis requires longitudinal data, most of which focuses
on \env{}.
Furthermore, we only expect to see deleterious synonymous
mutations at high frequencies if they are linked to repeated adaptive
substitions. Otherwise, they remain at low frequencies and can only be observed
by deep sequencing methods.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence data collection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Longitudinal intrapatient viral RNA sequences were collected from published
studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
show substantial population structure and were discarded (see supplementary
materials); a total of 11 patients with 4-23 time points each and approximately 10
sequences per time point were analyzed. The time intervals between two
consecutive sequences ranged from 1 to 34 months, most of them approximately 6 to 10 months.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimate of times to fixation/extinction, polymorphisms were
binned by frequency and the time to reaching the first boundary (fixation or
extinction) was stored. The fixation probability was determined as the long-time
limit of the resulting curves. Mutations that reached high frequency but neither
fixed nor got extinct were classified as ``floating'' and not analyzed further.

The SHAPE scores quantifying the degree of base pairing of individuals sites in
the HIV genome were downloaded from the journal
website~\citep{watts_architecture_2009}.
Where possible, SHAPE reactivities were assigned to sites in the multiple
sequence alignments for each patient through the alignment to the sequence
of the NL4.3 virus used in \citep{watts_architecture_2009}. Problematic
assignments in indel-rich regions were excluded from the analysis.

The variable loops and flanking regions were identified manually starting from
the annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
similar approach was used to label the C2-V5 region sequenced in
ref.~\citep{shankarappa_consistent_1999}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using
FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
\env{}-only simulations were performed and yielded comparable results. In each
simulation, four random parameters were drawn form prior distributions: (i)
fraction of deleterious synonymous sites (flat between 0.75 and 1.0); (ii)
deleterious effect at those sites (logarithmic between $10^{-4}$ and $10^{-2}$);
(iii) average beneficial effect of an escape mutation (logarithmic between
$10^{-2.5}$ and $10^{-1.5}$); (iv) rate of introduction of new epitopes
(logarithmic between $10^{-3}$ and $10^{-2}$ per generation). After these four
parameters were fixed, the simulation started. All the deleterious synonymous
sites had the same fitness effect (variations were tested and produced the same
results), and no epitopes (i.e. no escape mutations) were initially present.
After 30 generations of burn-in to create some genetic diversity, new epitopes
were introduced at a constant rate (as specified above). Each time such an
epitope was actually introduced, its fitness advantage was drawn from an
exponential distribution with fixed average value (as specified above).
For the models with competition within epitopes, a complex epistatic fitness
landscape was manually crafted, in which each single mutant provides the whole
beneficial effect of escape. For the models with time-dependent selection due to
the immune system catching up, the beneficial effect of an escape mutation was
reverted at random after it had appeared, according to its integrated allele
frequency (see supplementary materials for more details).

For each set of parameters, fixation probabilities and probabilities of
synonymous polymorphisms $P_\text{interm}$ were calculated as averages over
approximately 100 repetitions (with different random seeds).

The areas below or above the neutral fixation probability (diagonal line)
were estimated from the binned fixation probabilities using linear
interpolation between the bin centers. This measure is sufficiently precise for
our purposes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Methods availability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All analysis and computer simulation scripts, as well as the sequence alignments
used, are available for download.
